Solar-Electric and Gas Powered, Long-Endurance UAV Sizing via Geometric Programming

### 2017-01-26 gassolarpaper/gassolarpaper.tex
more edits; plot formatting; variable tweaking

 \usepackage{varioref}%  smart page, figure, table, and equation referencing
 \usepackage{wrapfig}%   wrap figures/tables in text (i.e., Di Vinci style)
 \usepackage{threeparttable}% tables with footnotes
 \usepackage{dcolumn}%   decimal-aligned tabular math columns
  \newcolumntype{d}{D{.}{.}{-1}}
 \usepackage{nomencl}%   nomenclature generation via makeindex
  \makeglossary
 \usepackage{amssymb,amsmath}
 \usepackage{subfigure}% subcaptions for subfigures
 \usepackage{subfigmat}% matrices of similar subfigures, aka small mulitples
 \usepackage{fancyvrb}%  extended verbatim environments
 \fvset{fontsize=\footnotesize,xleftmargin=2em}
 \usepackage{lettrine}%  dropped capital letter at beginning of paragraph
%  \usepackage[dvips]{dropping}% alternative dropped capital package
 \usepackage[colorlinks]{hyperref}%  hyperlinks [must be loaded after dropping]
 \usepackage{float}
 \usepackage{longtable,booktabs,tabularx}
 \restylefloat{table}
 \usepackage{graphicx}
 \usepackage{caption}
 \usepackage{siunitx}
 \usepackage{multicol}
 \usepackage{indentfirst}
 \usepackage[labelfont=bf]{caption}

@@

 \author{
  Michael Burton \thanks{Master's Candidate, Aeronautics and Astronautics Engineering, 77 Mass Ave, Cambridge MA, 02139, AIAA Student.}
  \ and Warren Hoburg\thanks{Assistant Professor, Aeronautical and Astronautical Engineering, 77 Mass Ave, Cambridge MA, 02139, AIAA Member.}\\
  {\normalsize\itshape
   Massachusetts Institute of Technology, Cambridge, 02139, USA}\\
 }

@@

BSFC \> brake specific fuel consumption [lb/hr/hp] \\

@@

$C_D$ \> drag coefficient \\

@@

$E_{\text{battery}}$ \> energy stored in battery [J] \\

@@

@@

$m_{\text{h}}$ \> modified horizontal tail lift coefficient \\

@@

$m_{\text{w}}$ \> modified wing lift coefficient \\

@@

$N_{\text{max}}$ \> load factor\\

@@

@@

$W_{\text{battery}}$ \> battery weight [N] \\

@@

@@

$\rho_{\text{carbon fiber}}$ \> density of carbon fiber [kg/m$^3$] \\

@@

$\sigma_{\text{carbon fiber}}$ \> maximum carbon fiber stress [Pa] \\

@@

Long-endurance, station-keeping aircraft are potential solutions to providing internet, communication, and persistent survellience. 
Recently, tech companies such as Google\cite{googletitan} and Facebook\cite{aquila} have explored the possibility of using solar-electric powered aircraft to provide internet to parts of the world where 4G or 3G network is not available. 
Engineering firms like Aurora Flight Sciences\cite{orion} and Vanailla Aircraft\cite{vanilla} have developed gas powered long-endurance platforms as a means of providing continual survellience for days at a time.  
However, sizing a long-endurance aircraft is not obvious because of the multifacted interaction between aerodynamics, structural weight, solar energy, wind speed, and other disciplines.
Additionally, because few long-endurance gas powered aircraft have been built and only Airbus's Zephyr\cite{zephyr} solar aircraft has been successfully flown, the operational limits of these aircraft is relatively unknown. 
Even more unclear is understanding the trade offs between gas powered and solar-electric powered archictures.

This paper presents a physics-based optimization model that uses geometric programming as systematic and rapid approach to explore this design space and highlight trade offs between the two proposed architecutres.
Geometric programming, a form of convex optimization, is chosen as a means of evaluating the design space because of its rapid solve time and guaranteed convergence to a global optimum.\cite{gp}
Models and equations representing the interaction of the various disciplines are expressed in a geometric programming form that are then combined to form an optimization model. 
The optimization is solved multiple times for different design parameters such as endurance, latitude, and percentile wind speed to observe how aircraft size and weight are affected by different requirements.

@@

The key driving requirement for a gas powered aircraft is endurance.  
As gas-powered aircraft are endurance limited by the amount of fuel that they can carry, longer endurance requires more fuel and hence a larger aircraft.  
Therefore, a trade study was done for the gas powered aircraft to determine how weight increases for longer endurance.
Because gas-powered aircraft are not affected by the solar flux, their station-keeping ability at different latitudes only depends on wind speed. 
If a gas-powered aircraft can fly at the latitude with the worst wind speed in can theoretically fly anywhere else in the world.

@@

For long-endurance gas powered aircraft, the endurance requirement is defined by amount of time the aircraft needs to remain in the air without refueling.  
For a solar-electric powered aircraft, however, if it can fly during the shortest day of the year, December 21st, then it can theortically fly at any time the year.  
Therefore, the endurance requirement for solar-electric powere aircaft is assumed to be a requirement to fly during the winter solstice.  
To achieve this design constraint the aircraft must be able to fly with limited solar energy, discussed in Section , and with higher wind speeds as shown in Figure~\ref{f:windvsmonth}.  
Because the endurance requirement for solar-electric powered aircraft is set by the winter solstice and because wind speeds are higher during winter months, only wind speed data from December will be used for the aircraft sizing and performance analysis presented in this paper.

@@

Gas powered aircraft generally fly between 5,000 to 25,000 ft due to payload requirements.  
Because of a naturally aspirated engine and linearly increasing wind speeds from 0 to 35,000 ft shown in Figure~\ref{f:altvswind} gas powered aircraft will generally be more efficiect at lower altitudes and generally fly at a given latitude to meet a payload requirement.  
For this design study, it is assumed that the aircraft is flying at 15,000 ft.

Solar-electric powered aircraft tend to fly around 60,000 ft to meet payload requirements, avoid cloud coverage and fly in lower wind speeds which have a minimum around 60,000 shown in Figure~\ref{f:altvswind}.  For the solar-electric powered aircraft altitude is optimized as explained in Section.

@@

Latitude additionally affects the solar-electric powered aircraft because at higher latitudes there is less daylight and therefore less solar energy during the winter months. This is further discussed in Section .

@@

For the purposes of this design study, a percentile wind speed corresponds to that percentile of wind speed data in December across all longitudes and spanning the years 2005-2015.  
To understand what a certain percentile wind speed implies operationally an autocorrelation analysis was done for multiple cites at varying latitudes at 60,000 during December. 
Boston and Rome during December of 2015 are shown in Figures~\ref{f:bostonwinds} and~\ref{f:romewinds}.

@@

The properties of a geometric program allow solution algorithms to guarantee convergence to a global optimum, provided that there exists a feasible solution to the set of constraint functions.  
Additionally, geometric programs can be solved rapidly with existing algorithms.  
Using state of the art, standard interior-point algorithms, GPs with 1,000 variables and 10,000 constraints converge to a solution in tens of seconds.\cite{gp}  
To solve an aircraft design problem, all physical models are expressed as constraints on the GP.\cite{hoburgthesis} 
A python package for defining and manipulating geometric programs, GPkit\cite{gpkitdocs}, is used to create the models described in this paper.  A commerical solver, mosek\cite{mosek}, is used to solve the GP. \\

@@

Because wind speeds also vary with latitude several constraints are needed to optimize altitude. 
Constraints relating wind speed and air density were generated using the data fitting technique described by Hoburg\cite{fitting}.
Air density was chosen as the dependent variable instead of altitude to eliminate an extra equation relating altitude to air density. 
Altitude can be post-calculated from the optimized air density. 
Because wind speeds are low and solar flux is higher below 20 degrees latitude no constraints for latitudes below 20 degrees are actually included in the optimization. 
Additionally, because long-endurance aircraft are likely to fly over land and the majority of land mass lies in the Northern Hemisphere, only wind speed data from the Northern Hemisphere is considered.
Thus, an aircraft optimized for $\pm$30 will have 11 different wind speed and solar flux constraints that must be satisfied, one for each degree of latitude. 
Therefore, 41 wind speed constraints were generated from 20-60 degrees North latitude. (See Appendix A) 
The constraint on wind speed and air density at the 35 degree latitude is

@@

At a given time of year DOY, and latitude $\phi$, there exists an availble solar irradiated energy $(E/S)_{\text{sun}}$ per unit area during one day

@@

To operate continually for multiple days the total available solar energy $(E/S)_{\text{sun}}$, must be greater than the required solar energy per unit area to power the aircraft during the day $(E/S)_{\text{day}}$ and the energy required to power the aircraft during the night via batteries $E_{\text{batt}}$,\cite{solartech}

\begin{align}
    \label{e:solarreq}
    (E/S)_{\text{sun}}  &\geq (E/S)_{\text{day}} + \frac{E_{\text{battery}}}{\eta_{\text{charge}}\eta_{\text{solar}} S_{\text{solar}}} \\
    \label{e:solarbatt}
    E_{\text{battery}} &\geq \frac{P_{\text{oper}}t_{\text{night}}}{\eta_{\text{discharge}}} + (E/S)_C \eta_{\text{solar}} S_{\text{solar}}
\end{align}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{lat30.pdf}
    \caption{\textbf{The available solar power per unit area during the winter solstice in the Northern Hemisphere at $\phi=30^{\circ}$. For a solar-electirc powered aircraft to be operational for more than 24 hours, the area under the red curve must be greater than the areas shown in blue, magenta and green.}}
	\label{f:lat30}
	\end{center}
\end{figure}

@@

For simplicity, energy and power will refer to energy and power per unit area throughout the rest of this section (i.e. $(P/S) = P$). 
The solar irradiated power $P_{\text{sun}}$ a function of $\theta$, the angle between the normal to the flat surface, or aircraft wing, and the sun beam.\cite{solar}

@@

    Using a trapezoidal integration of Equation~\eqref{e:solares}, the total available solar energy per unit area can be obtained for a given latitude and day of the year. Because only Equations~\eqref{e:solarreq}-\eqref{e:solarssolar} are GP compatible, the total solar energy per unit area per day, the length of the day and the length of the night is calculated from the latitude and the day of the year prior to an optimization solve.

@@

This verison comes from assuming that $\text{BSFC}$ and the power to weight ratio, $(P_{\text{shaft}}/W)$, are constant during the considered flight segment. 
One way to obtain a constant power to weight ratio is a constant velocity and constant lift coefficient.\cite{br2}
$W_{\text{ave}}$ is assumed to be the geometric mean, defined as

@@

    To make Equation~\eqref{e:be1} GP compatible, a Taylor expansion of the exponentiated right hand side is used,\cite{hoburgthesis}

@@

    Equations~\eqref{e:brzbre} and~\eqref{e:brtaylor} are monomial and posynomial respectively and therefore GP compatible. \\

    For long-endurance aircraft, missions can last days, causing the power to weight ratio $(P_{\text{shaft}}/W)$ to vary significantly during the course of the flight.  
    Equations~\eqref{e:brzbre},~\eqref{e:brtaylor}, and~\eqref{e:slfweight} can be discretized to account for this.

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{JH01polar.pdf}
    \caption{\textbf{The posynomial fit, Equation~\eqref{e:aerodragprof}, of the \emph{JH01} airfoil drag polar is represented as a solid line.  The open circles are the calculated drag polars from XFOIL. (Log space RMS error = 0.00489.)}}
	\label{f:JH01polar}
	\end{center}
\end{figure}

As a preliminary sizing study the aspect ratio is assumed to be a constant, $A = 30$.  A more detailed sizing study with a wing structural model, discussed later, will allow the aspect ratio to be a free variable.

\begin{equation}
    \label{e:weightmtow}
    \text{MTOW} \geq W_{\text{structural}}  + W_{\text{payload}} + W_{\text{fuel}} + W_{\text{engine}}
\end{equation}

A similiar weight breakdown is done for the solar-electric powered aircraft,

\begin{align}
    \label{e:weightsmtow}
    \text{MTOW} &\geq W_{\text{structural}} + W_{\text{payload}} + W_{\text{solar}} + W_{\text{battery}} + W_{\text{motor}} \\
    W_{\text{solar}} &\geq \rho_{\text{solar}} S_{\text{solar}} g \\
    W_{\text{battery}} &\geq \frac{E_{\text{battery}}}{h_{\text{battery}}} g
\end{align}

@@

% %                      &        & $h_{\text{battery}}$ [Whr/kg]  & 350 \\

@@

A wing spar is one of the main structural elements of both aircraft. 
A conservative approach to calculating the size of the spar is to assume that the spar carries all of the bending loads caused by the lifting loads and weight of the aircraft.  
It is assumed that there are two out of plane bending cases to which both the gas powered and the solar-electric powered aircraft are subject: steady level flight bending case and gust load case.

@@

Because long-endurance aircraft typically have high aspect ratio wing they are considered flexible aircraft. 
To account for this flexiblity and size a structural wing spar that will ensure enough rigidity in the event of a gust load, a distributed gust lifting load is added to the steady level flight loading distribution in Equation~\ref{e:qbar}

@@

where $\bar{c}(y)$ is pre-computed prior to the optimization solve for a taper ratio $\lambda = 0.5$. The weight of the wing for the gas powered aircraft is assumed to be the weight of the spar cap plus the weight of the wing

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{arctanfit.pdf}
    \caption{\textbf{Comparison of monomial approximation of $f(x) = \arctan{x}$ to real function, $x \in [0,0.7]$. (Log space RMS error = 0.044)}}
	\label{f:arctanfit}
	\end{center}
\end{figure}

@@

% \begin{figure}[H]
% 	\begin{center}
% 	\includegraphics[width=0.6\textwidth]{gustvschord.pdf}
%     \caption{ \textbf{ Graphical representation of assumption that the local chord is equal to the local lift generated from the elliptical distribution. Assumes the center to wing weight ratio, $\frac{W_{\text{cent}}}{W_{wing}} = 0.14$. }}
% 	\label{f:gustvschord}
% 	\end{center}
% \end{figure}

@@

\begin{equation}
    \label{e:moispar}
    I = \frac{w_{\text{cap}}t_{\text{cap}}^3}{6} 2w_{\text{cap}}t_{\text{cap}}\left( \frac{h_{\text{cap}}}{2} + \frac{t_{\text{cap}}}{2} \right)^2
\end{equation}

@@

\begin{equation}
    \label{e:moispar}
    I \leq 2w_{\text{cap}}t_{\text{cap}}\left(\frac{h_{\text{cap}}}{2}\right)^2
\end{equation}

There are also geometric constraints imposed on the width and thickness.  The total spar cap thickness cannot be greater than the thickness of the airfoil cross section, $\tau_t = 0.115$.  The width of the spar cap is assumed by no greater than 30\% of chord, $\tau_w = 0.3$.

@@

The wing spar at the root must be strong enough to withstand the bending root moment and stiff enough to not exceed some deflection limit.  Both constraints are imposed in the optimization model as

\begin{align}
    \label{e:stresscont}
    \sigma_{\text{carbon fiber}} &\geq \mathcal{M}_0 \frac{h_{\text{cap}}+t_{\text{cap}}}{I}\\
    \label{e:defcont}
    w_n &\leq w_{\text{max}}.
\end{align}

The maximum stress for carbon fiber is, $\sigma_{\text{carbon fiber}} = 475$ [MPa].\cite{carbonfiber}
The tip deflection is constrained to be less than 20\% of the half span, $\frac{w_{\text{max}}}{b/2} = 0.2$.

@@

\begin{align}
    \label{e:sparmass}
    \Delta W_i &\geq \rho_{\text{carbon fiber}} w_{\text{cap}_i}t_{\text{cap}_i} \frac{b/2}{n-1}g \\
    \label{e:sparmasssum}
    W_{\text{spar}} &\geq 2 \sum\limits_{1}^{n-1} \Delta W_i
\end{align}

where $\rho_{\text{carbon fiber}} = 1.4$ [g/cm$^3$].\cite{carbonfiber}

@@

\begin{equation}
    \label{e:wingskinweight}
    W_{\text{skin}} \geq 2 \rho_{\text{carbon fiber}} S g 
\end{equation}

where $\rho_{\text{carbon fiber}} = 0.043$ [g/cm$^2$].\cite{carbonfiber}

@@

The tail boom has a diameter $d$, root wall thickness $t_0$, root moment of inertia $I_0$, modulus $E$, and density $\rho_{\text{carbon fiber}} = 1.4$ [g/cm$^3$], and length $l_{\text{h}}$. 
The total mass and root bending inertia are imposed in the optimization model as

\begin{align}
    m &\geq \pi \rho_{\text{carbon fiber}} t_0 d l_{\text{h}} \left( 1 + \frac{1}{2} k\right) \\
    I_0 &\leq \pi t_0 d^3/8
\end{align}

where the index $k=0$ corresponds to a uniform wall thickness and stiffness, and $k=1$ corresponds to a linear drop-off to zero.  For both the solar-electric and gas powered aircraft $k=0.8$.  When the tail boom is loaded at the endpoint $x=l_{\text{h}}$, by the horizontal tail lift $L_{\text{h}}$, the end deflection angle follows from standard beam analysis.

\begin{align}
    \label{e:boomdefl}
    \theta &\geq \frac{L_{\text{h}} l_{\text{h}}^2}{EI_0} \frac{1+k}{2} \\
    L_{\text{h}} &= \frac{1}{2} C_{L_{\text{h}}} \rho V^2 S_{\text{h}}
\end{align}

The horizontal tail is sized to satisfy a horizontal tail volume coefficient condition, $V_{\text{h}} = 0.45$.\cite{aircraftrules}

@@

The vertical tail is sized to meet a conservative volume tail coefficient, $V_{\text{v}}= 0.02$,\cite{aircraftrules}

@@

where $l_{\text{v}}$ is the vertical tail moment arm and assume to be equal to the horizontal tail moment arm, $l_{\text{v}} = l_{\text{h}}$.

Both the horizontal and vertical tails are assumed to have a carbon fiber skin and solid foam interior where their respective densities are $\rho_{\text{carbon fiber}} = 0.1$ [g/cm$^2$], $\rho_{\text{foam}} = 1.5$ [lbf/ft$^3$]. 
The weight of the horizontal and vertical tails is

>>>>

 \usepackage{varioref}%  smart page, figure, table, and equation referencing
 \usepackage{wrapfig}%   wrap figures/tables in text (i.e., Di Vinci style)
 \usepackage{threeparttable}% tables with footnotes
 \usepackage{dcolumn}%   decimal-aligned tabular math columns
  \newcolumntype{d}{D{.}{.}{-1}}
 \usepackage{nomencl}%   nomenclature generation via makeindex
  \makeglossary
 \usepackage{amssymb,amsmath}
 \usepackage{subfigure}% subcaptions for subfigures
 \usepackage{subfigmat}% matrices of similar subfigures, aka small mulitples
 \usepackage{fancyvrb}%  extended verbatim environments
 \fvset{fontsize=\footnotesize,xleftmargin=2em}
 \usepackage{lettrine}%  dropped capital letter at beginning of paragraph
%  \usepackage[dvips]{dropping}% alternative dropped capital package
 \usepackage[colorlinks]{hyperref}%  hyperlinks [must be loaded after dropping]
 \usepackage{float}
 \usepackage{longtable,booktabs,tabularx}
 \restylefloat{table}
 \usepackage{graphicx}
 \usepackage{caption}
 \usepackage{siunitx}
 \usepackage{multicol}
 \usepackage{indentfirst}
 \usepackage[labelfont=bf]{caption}
 \usepackage{multirow}

@@

 \author{
  Michael Burton \thanks{Master's Candidate, Aeronautics and Astronautics Engineering, 77 Mass Ave, Cambridge MA, 02139, AIAA Student.}
  \ and Warren Hoburg\thanks{Assistant Professor, Aeronautics and Astronautics Engineering, 77 Mass Ave, Cambridge MA, 02139, AIAA Member.}\\
  {\normalsize\itshape
   Massachusetts Institute of Technology, Cambridge, 02139, USA}\\
 }

@@

BSFC \> brake specific fuel consumption [kg/kW/hr] \\

@@

$C_D$ \> aircraft drag coefficient \\

@@

$E_{\text{batt}}$ \> energy stored in battery [J] \\

@@

$(E/S)_{\text{C}}$ \> low irradiance solar energy per unit area [Whr/m$^2$] \\
$(E/S)_{\text{day}}$ \> daily operational solar energy per unit area [Whr/m$^2$] \\

@@

$m_{\text{fac-wing}}$ \> margin factor\\
$m_{\text{fac-wing}}$ \> wing weight margin factor \\

@@

@@

$N_{\text{max}}$ \> safety load factor\\

@@

$(P/S)_{\text{min}}$ \> minium operational solar power \\
\>per unit area [Whr/m$^2$] \\

@@

$W_{\text{batt}}$ \> battery weight [N] \\

@@

$\eta_{\text{motor}}$ \> electric motor efficiency \\

@@

$\rho_{A_{\text{cfrp}}}$ \> area density of carbon fiber [kg/cm$^2$] \\
$\rho_{\text{cfrp}}$ \> density of carbon fiber [kg/cm$^3$] \\

@@

$\sigma_{\text{cfrp}}$ \> maximum carbon fiber stress [Pa] \\

@@

Long-endurance, station-keeping aircraft are potential solutions to providing internet, communication, and persistent surveillance. 
Recently, tech companies such as Google\cite{googletitan} and Facebook\cite{aquila} have explored the possibility of using solar-electric powered aircraft to provide internet to parts of the world where 4G or 3G network is not available. 
Engineering firms like Aurora Flight Sciences\cite{orion} and Vanilla Aircraft\cite{vanilla} have developed gas powered long-endurance platforms as a means of providing continual surveillance for days at a time.  
However, sizing a long-endurance aircraft is not obvious because of the multifaceted interaction between aerodynamics, structural weight, solar energy, wind speed, and other disciplines.
Additionally, because few long-endurance gas powered aircraft have been built and only Airbus has successfully built and tested a solar aircraft for multiple days\cite{zephyr}, the operational limits of these aircraft is relatively unknown. 
Even more unclear are the trade offs between gas powered and solar-electric powered architectures.

This paper presents a physics-based optimization model that uses geometric programming as a systematic and rapid approach to explore this design space and highlight trade offs between the two proposed architectures.
Geometric programming, a form of convex optimization, is chosen as a means of evaluating the design space because of its rapid solve time and guaranteed convergence to a global optimum.\cite{gp}
Models and equations representing the interaction of the various disciplines are expressed in a geometric programming form that are then combined to form an optimization model. 
The optimization is solved multiple times for different design parameters such as endurance, latitude, and percentile wind speed to observe how aircraft size and weight are affected by different requirements.

@@

The key driving requirement for a gas powered aircraft is endurance.  
As gas-powered aircraft are endurance limited by the amount of fuel that they can carry, longer endurance requires more fuel and hence a larger aircraft.  
Because gas-powered aircraft are not affected by the solar flux, their station-keeping ability at different latitudes only depends on the local wind speed. 
If a gas-powered aircraft can fly at the latitude with the worst wind speed, in can theoretically fly anywhere else in the world.

@@

For long-endurance gas powered aircraft, the endurance requirement is defined by amount of time the aircraft needs to remain in the air without refueling.  
For a solar-electric powered aircraft, however, if it can fly during the shortest day of the year, December 21st, then it can thoretically fly at any time the year.  
Therefore, the endurance requirement for solar-electric powered aircraft is assumed to be a requirement to fly during the winter solstice.  
To achieve this design constraint the aircraft must be able to fly with limited solar energy, discussed in Section IV.B, and with higher wind speeds as shown in Figure~\ref{f:windvsmonth}.  
Because the endurance requirement for solar-electric powered aircraft is set by the winter solstice and because wind speeds are higher during winter months, only wind speed data from December will be used for the aircraft sizing and performance analysis presented in this paper.

@@

Gas powered aircraft generally fly between 5,000 to 25,000 ft due to payload requirements.  
Because of a naturally aspirated engine and linearly increasing wind speeds from 0 to 35,000 ft shown in Figure~\ref{f:altvswind} gas powered aircraft will generally be more efficient at lower altitudes and generally fly at a given latitude to meet a payload requirement.

Solar-electric powered aircraft tend to fly around 60,000 ft to meet payload requirements, avoid cloud coverage and fly in lower wind speeds which have a minimum around 60,000 shown in Figure~\ref{f:altvswind}.  For the solar-electric powered aircraft altitude is optimized as explained in Section IV.A.

@@

Latitude additionally affects the solar-electric powered aircraft because at higher latitudes there is less daylight and therefore less solar energy during the winter months. This is further discussed in Section IV.B.

@@

For the purposes of this design study, a percentile wind speed corresponds to that percentile of the wind speed data in December, across all longitudes and spanning the years 2005-2015.  
To understand what a given percentile wind speed implies operationally an autocorrelation analysis was done for multiple cites at varying latitudes at 60,000 during December. 
Wind speeds 60,000 ft over Boston and Rome during December of 2015 are shown in Figures~\ref{f:bostonwinds} and~\ref{f:romewinds}.

@@

The properties of a geometric program allow solution algorithms to guarantee convergence to a global optimum, provided that there exists a feasible solution to the set of constraint functions.  
Additionally, geometric programs can be solved rapidly with existing algorithms.  
Using state of the art, standard interior-point algorithms, GPs with 1,000 variables and 10,000 constraints converge to a solution in tens of seconds.\cite{gp}  
To solve an aircraft design problem, all physical models are expressed as constraints on the GP.\cite{hoburgthesis} 
A python package for defining and manipulating geometric programs, GPkit\cite{gpkitdocs}, is used to create the models described in this paper.  A commercial solver, mosek\cite{mosek}, is used to solve the GP. \\

@@

Because wind speeds also vary with latitude several constraints are needed to optimize altitude. 
Constraints relating wind speed and air density were generated using the data fitting technique described by Hoburg\cite{fitting}.
Air density was chosen as the dependent variable instead of altitude to eliminate an extra equation relating altitude to air density. 
Altitude can be post-calculated from the optimized air density. 
Because wind speeds are low and solar flux is higher below 20 degrees latitude, no constraints for latitudes below 20 degrees are actually included in the optimization. 
Additionally, because long-endurance aircraft are likely to fly over land and the majority of land mass lies in the Northern Hemisphere, only wind speed data from the Northern Hemisphere is considered.
Thus, an aircraft optimized for $\pm$30 will have 11 different wind speed and solar flux constraints that must be satisfied, one for each degree of latitude. 
Therefore, 41 wind speed constraints were generated from 20-60 degrees North latitude. (See Appendix A) 
The constraint on wind speed and air density at the 35 degree latitude is

@@

At a given time of year DOY, and latitude $\phi$, there exists an available solar irradiated energy $(E/S)_{\text{sun}}$ per unit area during one day

@@

To operate continually for multiple days the total available solar energy $(E/S)_{\text{sun}}$, must be greater than the required solar energy per unit area to power the aircraft during the day $(E/S)_{\text{day}}$, and the energy required to power the aircraft during the night via batteries $E_{\text{batt}}$,\cite{solartech}

\begin{align}
    \label{e:solarreq}
    (E/S)_{\text{sun}}  &\geq (E/S)_{\text{day}} + \frac{E_{\text{batt}}}{\eta_{\text{charge}}\eta_{\text{solar}} S_{\text{solar}}} \\
    \label{e:solarbatt}
    E_{\text{batt}} &\geq \frac{P_{\text{oper}}t_{\text{night}}}{\eta_{\text{discharge}}} + (E/S)_C \eta_{\text{solar}} S_{\text{solar}}
\end{align}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.7\textwidth]{lat30.pdf}
    \caption{\textbf{The available solar power per unit area during the winter solstice in the Northern Hemisphere at $\phi=30^{\circ}$. For a solar-electric powered aircraft to be operational for more than 24 hours, the area under the red curve must be greater than the areas shown in blue, magenta and green.}}
	\label{f:lat30}
	\end{center}
\end{figure}

@@

For simplicity, energy and power will refer to energy and power per unit area throughout the rest of this section (i.e. $(P/S) = P$). 
The solar irradiated power $P_{\text{sun}}$, is a function of $\theta$, the angle between the normal to the flat surface, or aircraft wing, and the sun beam.\cite{solar}

@@

    Using a trapezoidal integration of Equation~\eqref{e:solares}, the total available solar energy per unit area can be obtained for a given latitude and day of the year. Because only Equations~\eqref{e:solarreq}-\eqref{e:solarssolar} are GP compatible, the total solar energy per unit area per day, the length of the day and the length of the night are calculated from the latitude and the day of the year prior to an optimization solve.

@@

This version comes from assuming that $\text{BSFC}$ and the power to weight ratio, $(P_{\text{shaft}}/W)$, are constant during the considered flight segment. 
One way to obtain a constant power to weight ratio is a constant velocity and constant lift coefficient.\cite{br2}
$W_{\text{ave}}$ is assumed to be the geometric mean, defined as

@@

    To make Equation~\eqref{e:breguetendurance} GP compatible, a Taylor expansion of the exponentiated right hand side is used,\cite{hoburgthesis}

@@

    Equations~\eqref{e:brzbre} and~\eqref{e:brtaylor} are monomial and posynomial respectively and therefore GP compatible. For long-endurance aircraft, missions can last days, causing the power to weight ratio $(P_{\text{shaft}}/W)$ to vary significantly during the course of the flight.  
    Equations~\eqref{e:brzbre},~\eqref{e:brtaylor}, and~\eqref{e:slfweight} can be discretized to account for this.

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{JH01polar.pdf}
    \caption{\textbf{The posynomial fit, Equation~\eqref{e:aerodragprof}, of the \emph{JH01} airfoil drag polar is represented as a solid line.  The open circles are the calculated drag polars from XFOIL. (Log space RMS error $= 0.00489$.)}}
	\label{f:JH01polar}
	\end{center}
\end{figure}

@@

% %                      &        & $h_{\text{batt}}$ [Whr/kg]  & 350 \\

@@

A wing spar is one of the main structural elements of both aircraft. 
A conservative approach to calculating the size of the spar is to assume that the spar carries all of the bending loads caused by the lifting loads and weight of the aircraft.  
It is assumed that there are two out of plane bending cases to which both the gas powered and the solar-electric powered aircraft are subject: steady level flight case and gust loads.

@@

Because long-endurance aircraft typically have high aspect ratio wing they are considered flexible aircraft. 
To account for this flexibility and size a structural wing spar that will ensure enough rigidity in the event of a gust load, a distributed gust lifting load is added to the steady level flight loading distribution in Equation~\ref{e:qbar}

@@

where $\bar{c}(y)$ is pre-computed prior to the optimization solve for a taper ratio $\lambda = 0.5$. The safety load factor is $N_{\text{max}}=2$ for the gust loading case. The weight of the wing for the gas powered aircraft is assumed to be the weight of the spar cap plus the weight of the wing

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{arctanfit.pdf}
    \caption{\textbf{Comparison of monomial approximation of $f(x) = \arctan{x}$ to real function, $x \in [0,0.7]$. (Log space RMS error = 0.0246)}}
	\label{f:arctanfit}
	\end{center}
\end{figure}

@@

@@

\begin{equation}
    \label{e:moispar}
    I = \frac{w_{\text{cap}}t_{\text{cap}}^3}{6} + 2w_{\text{cap}}t_{\text{cap}}\left( \frac{h_{\text{cap}}}{2} + \frac{t_{\text{cap}}}{2} \right)^2.
\end{equation}

@@

\begin{equation}
    \label{e:moispar}
    I \leq 2w_{\text{cap}}t_{\text{cap}}\left(\frac{h_{\text{cap}}}{2}\right)^2.
\end{equation}

There are also geometric constraints imposed on the width and thickness.  The total spar cap thickness cannot be greater than the thickness of the airfoil cross section, $\tau_t = 0.115$.  The width of the spar cap is assumed no greater than 30\% of chord, $\tau_w = 0.3$.

@@

The wing spar at each section root must be strong enough to withstand the bending moment and stiff enough to not exceed some deflection limit.  Both constraints are imposed in the optimization model as

\begin{align}
    \label{e:stresscont}
    \sigma_{\text{cfrp}} &\geq \mathcal{M}_i \frac{h_{\text{cap}_i}+t_{\text{cap}_i}}{I_i}\\
    \label{e:defcont}
    w_n &\leq w_{\text{max}}.
\end{align}

The ultimate compressive strength for carbon fiber is, $\sigma_{\text{cfrp}} = 570$ [MPa].\cite{carbonfiber}
The tip deflection is constrained to be less than 20\% of the half span, $\frac{w_{\text{max}}}{b/2} = 0.2$.

@@

\begin{align}
    \label{e:sparmass}
    \Delta W_i &\geq \rho_{\text{cfrp}} w_{\text{cap}_i}t_{\text{cap}_i} \frac{b/2}{n-1}g \\
    \label{e:sparmasssum}
    W_{\text{spar}} &\geq 2 \sum\limits_{1}^{n-1} \Delta W_i
\end{align}

where $\rho_{\text{cfrp}} = 1.6$ [g/cm$^3$].\cite{carbonfiber}

@@

\begin{equation}
    \label{e:wingskinweight}
    W_{\text{skin}} \geq 2 \rho_{A_{\text{cfrp}}} S g 
\end{equation}

where $\rho_{A_{\text{cfrp}}} = 0.049$ [g/cm$^2$], or the area density of one ply of carbon fiber.\cite{carbonfiber}

@@

The tail boom has a diameter $d$, root wall thickness $t_0$, root moment of inertia $I_0$, modulus $E$, and density $\rho_{\text{cfrp}} = 1.6$ [g/cm$^3$], and length $l_{\text{h}}$. 
The total mass and root bending inertia are imposed in the optimization model as

\begin{align}
    m &\geq \pi \rho_{\text{cfrp}} t_0 d l_{\text{h}} \left( 1 + \frac{1}{2} k\right) \\
    I_0 &\leq \pi t_0 d^3/8
\end{align}

where the index $k=0$ corresponds to a uniform wall thickness and stiffness, and $k=1$ corresponds to a linear drop-off to zero.  For both the solar-electric and gas powered aircraft $k=0.8$.  
When the tail boom is loaded at the endpoint $x=l_{\text{h}}$, by the horizontal tail lift $L_{\text{h}}$, the end deflection angle follows from standard beam analysis

\begin{align}
    \label{e:boomdefl}
    \theta &\geq \frac{L_{\text{h}} l_{\text{h}}^2}{EI_0} \frac{1+k}{2} \\
    L_{\text{h}} &= \frac{1}{2} C_{L_{\text{h}}} \rho V^2 S_{\text{h}}.
\end{align}

The horizontal tail is sized to satisfy a horizontal tail volume coefficient condition, $V_{\text{h}} = 0.45$,\cite{aircraftrules}

@@

The vertical tail is sized to meet a conservative tail volume coefficient, $V_{\text{v}}= 0.04$,\cite{aircraftrules}

@@

where $l_{\text{v}}$ is the vertical tail moment arm, assumed to be equal to the horizontal tail moment arm, $l_{\text{v}} = l_{\text{h}}$.

Both the horizontal and vertical tails are assumed to have a carbon fiber skin and solid foam interior where their respective densities are $\rho_{A_{\text{cfrp}}} = 0.049$ [g/cm$^2$], $\rho_{\text{foam}} = 1.5$ [lbf/ft$^3$]. 
The weight of the horizontal and vertical tails is

\begin{align}
    \label{e:htweight}
    W_{\text{h}}/m_{fac} &= \rho_{\text{foam}} \frac{S_{\text{h}}^2}{b_{\text{h}}} \bar{A} + g\rho_{A_{\text{cfrp}}} S_{\text{h}} \\
    \label{e:vtweight}
    W_{\text{v}}/m_{fac} &= \rho_{\text{foam}} \frac{S_{\text{v}}^2}{b_{\text{v}}} \bar{A} + g\rho_{A_{\text{cfrp}}} S_{\text{v}}
\end{align}
